---
title: "Project specific filtering"
output: rmarkdown::html_vignette
params:
  group:
    label: "Grouping variable"
    value: broad_histology
    input: string
  numCaller:
    label: "Least Number of callers to have called fusion"
    value: 2
    input: integer
  numSample:
    label: "Least Number of samples to have fusion per group"
    value: 2
    input: integer
  numGroup:
    label: "Max number of groups found in"
    value: 1
    input: integer
  limitMultiFused:
    label: "Max number of times gene can be fused per sample"
    value: 5
    input: integer  
  cohort:
    label: "Boolean for analysis is cohort"
    value: TRUE
    input: boolean     


vignette: >
  %\VignetteIndexEntry{Project specific filtering}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
    
---

K S Gaonkar

Filtered Fusions:
1. In-frame fusions is called in atleast 2 samples per histology OR
2. In-frame fusions is called in atleast 2 callers 
AND
Filtered-fusions found in more than 1 histology OR
Filtered-fusion doesn't have multi-fused gene (more than 5 times in sample)

Putative Driver:
Filtering for general cancer specific genes
Fusions with genes in either onco

This notebook assumes you are in OpenPBTA-analysis project folder structure.


```{r}
#load required packages
suppressPackageStartupMessages(library("readr"))
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("reshape2"))
suppressPackageStartupMessages(library("qdapRegex"))
library("annoFuse")

#read filtFusion files
strandedQCGeneFiltered_filtFusion<-readRDS(system.file("extdata", "standardFusionStrandedExp_QC_expression_GTExComparison_annotated.RDS", package = "annoFuse"))
polyaQCGeneFiltered_filtFusion<-readRDS(system.file("extdata", "standardFusionPolyaExp_QC_expression_GTExComparison_annotated.RDS", package = "annoFuse"))

QCGeneFiltered_filtFusion<-rbind(strandedQCGeneFiltered_filtFusion,polyaQCGeneFiltered_filtFusion)

# subset for recurrent fusion detection and multifused genes QC
fusion_calls<-unique(QCGeneFiltered_filtFusion)
# remove distance from intergenic fusions
fusion_calls$FusionName<-unlist(lapply(fusion_calls$FusionName,function(x) rm_between(x, "(", ")", extract = F)))


# get grouping column id
group<-params$group

# get histology file
clinical<-read.delim(system.file("extdata", "pbta-histologies.tsv", package = "annoFuse"),stringsAsFactors = FALSE)

# Least number of callers
numCaller<-params$numCaller

# Least number of samples per group
numSample<-params$numSample

# Max number of groups 
numGroup<-params$numGroup

# Max number of times gene can be fused per sample
limitMultiFused<-params$limitMultiFused

# TRUE/FALSE if cohort
cohort<-params$cohort

print("Raw calls from STARfusion and Arriba for PBTA")
table(fusion_calls$Caller)
```


```{r}
# aggregate caller and remove any fusion with LOCAL_REARRANGEMENT|LOCAL_INVERSION in annots column
fusion_calls<-aggregate_fusion_calls(fusion_calls,removeother=TRUE,filterAnnots="LOCAL_REARRANGEMENT|LOCAL_INVERSION")

# merge with histology file to use "broad_histology as group for cohort level filtering
fusion_calls<-merge(fusion_calls,clinical,by.x="Sample",by.y="Kids_First_Biospecimen_ID")

```


```{r}
# Gene fusion should be in-frame
# AND
#
# 1. Called by at least n callers
fusion_calls.summary <- called_by_n_callers(fusion_calls,numCaller)

# OR
# 2. Found in at least n samples in each group
sample.count <- samplecount_fusion_calls(fusion_calls,numSample,group)


```


```{r}
#filter QCGeneFiltered_filtFusion to keep recurrent fusions from above sample.count and fusion_calls.summary
QCGeneFiltered_recFusion<-fusion_calls %>%
  dplyr::filter(FusionName %in% unique(c(sample.count$FusionName,fusion_calls.summary$FusionName)))

```


```{r}

if (cohort){
# remove fusions that are in > numGroup
group.count <- groupcount_fusion_calls(fusion_calls ,group,numGroup)

# remove multi-fused genes
fusion_recurrent5_per_sample <- fusion_multifused(fusion_calls ,limitMultiFused)

# filter QCGeneFiltered_recFusion to remove fusions found in more than 1 group and multifused gene per samples
QCGeneFiltered_recFusionUniq<-QCGeneFiltered_recFusion %>%
  dplyr::filter(!FusionName %in% group.count$FusionName) %>%
  dplyr::filter(!Gene1A %in% fusion_recurrent5_per_sample$GeneSymbol |
                  !Gene2A %in% fusion_recurrent5_per_sample$GeneSymbol |
                  !Gene1B %in% fusion_recurrent5_per_sample$GeneSymbol | 
                  !Gene2B %in% fusion_recurrent5_per_sample$GeneSymbol)
} else {
# filter QCGeneFiltered_recFusion to remove fusion with multifused gene per samples
QCGeneFiltered_recFusionUniq<-QCGeneFiltered_recFusion %>%
  dplyr::filter(!Gene1A %in% fusion_recurrent5_per_sample$GeneSymbol |
                  !Gene2A %in% fusion_recurrent5_per_sample$GeneSymbol |
                  !Gene1B %in% fusion_recurrent5_per_sample$GeneSymbol | 
                  !Gene2B %in% fusion_recurrent5_per_sample$GeneSymbol)

}

```


```{r}
# filter for putative driver genes and mutifused genes per sample
putative_driver_fusions <-fusion_driver(standardFusioncalls=fusion_calls,annotated=TRUE)

# inframe putative driver fusion list
putative_driver_inframefusions<-putative_driver_fusions %>%
  dplyr::filter(Fusion_Type=="in-frame")
```


```{r}

fusion_calls<-rbind(putative_driver_fusions,QCGeneFiltered_recFusionUniq)
plotSummary(fusion_calls,groupby = "broad_histology",outputpdffile = "output.pdf")

```








