---
title: >
  Putative Driver filtering
author:
- name: Krutika Gaonkar
  affiliation: 
  - &id1 Center for Data-Driven Discovery in Biomedicine, Children’s Hospital of Philadelphia, USA
  email: gaonkark@chop.email.edu
- name: Federico Marini
  affiliation: 
  - &id2 Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  - Center for Thrombosis and Hemostasis (CTH), Mainz
  email: marinif@uni-mainz.de
date: "`r Sys.Date()`"
package: "`r paste('annoFuse', packageVersion('annoFuse'))"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Putative driver filtering}
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{annoFuse}
  %\VignetteKeywords{Software, Transcription, Visualization, GeneExpression, Annotation,
    GeneFusionDetection, GUI, ReportWriting}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


In this vignette we demonstrate using annoFuse, to filter putative oncogene fusion from a filtered fusion list (filtered fusion list `FilteredFusionAnnoFuse.tsv` was obtained for v16 OpenPBTA files : fusion_standardization -> fusion_filtering_QC -> annotate_fusion_calls ) and also plot recurrent fusion,recurrently fused genes and summary plot.

These high quality fusion calls are considered known and putative oncogenic driver fusions if the fusion is previously reported in TCGA or fusions containing gene partners that are known oncogenes, tumor suppressor genes, COSMIC genes, or transcription factors. 

Putative Driver Fusions that are found in more than 4 histologies were removed from further processing as these fusions are considered artifact or commonly occuring from manual review.
Or are recurrent fusions called by both callers and unique to specific histology in this filtered dataset. If fusions are from non-onocgenic genes that are fused more than 5 times per sample we consider these are artifacts and remove them from being further processed  



```{r}
# load required packages
suppressPackageStartupMessages(library("readr"))
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("reshape2"))
suppressPackageStartupMessages(library("qdapRegex"))

fusion_calls <- read_tsv(system.file("extdata", "FilteredFusionAnnoFuse.tsv", package = "annoFuse"))

# distance are being removed here to capture all intergenic fusions in count
# distance within () were making them count as unique instead of the same fusion
fusion_calls$FusionName <- unlist(lapply(fusion_calls$FusionName, function(x) rm_between(x, "(", ")", extract = F)))

print("Raw calls from STARfusion and Arriba for PBTA")
table(fusion_calls$Caller)
```


```{r}
# aggreate caller to unique fusion calls per Sample
fusion_calls <- aggregate_fusion_calls(fusion_calls, removeother = FALSE)

# gene list
geneListReferenceDataTab <- read.delim(
  system.file("extdata", "genelistreference.txt", package = "annoFuse"),
  stringsAsFactors = FALSE
)

# fusion list
fusionReferenceDataTab <- read.delim(
  system.file("extdata", "fusionreference.txt", package = "annoFuse"),
  stringsAsFactors = FALSE
)

# filter for driver fusions
putative_driver_fusions <- fusion_driver(standardFusioncalls = fusion_calls, annotated = TRUE, geneListReferenceDataTab = geneListReferenceDataTab, fusionReferenceDataTab = fusionReferenceDataTab)
```


```{r}

# checking if fusions are called in multiple histologies
# this will suggest these fusion calls are artifact or
# commonly occuring
found_in_morethan_4 <- groupcount_fusion_calls(putative_driver_fusions, group = "broad_histology", numGroup = 4) %>% arrange(desc(group.ct))

found_in_morethan_4
```

```{r}
# To scavenge back non-oncogenic recurrent/unique per broad histology fusions
scavenge_recurrent_non_onco <- fusion_calls <- aggregate_fusion_calls(fusion_calls,
  removeother = TRUE,
  filterAnnots = "LOCAL_REARRANGEMENT|LOCAL_INVERSION"
)

# Keep
# 1. Called by at least n callers
fusion_calls.summary <- called_by_n_callers(fusion_calls, numCaller = 2)

# OR
# 2. Found in at least n samples in each group
sample.count <- samplecount_fusion_calls(fusion_calls, numSample = 2, group = "broad_histology")

# Remove
# 1. non-oncogenic fusions that are in > numGroup
group.count <- groupcount_fusion_calls(fusion_calls, group = "broad_histology", numGroup = 1)

# 2. non-oncogenic multi-fused genes
fusion_recurrent5_per_sample <- fusion_multifused(fusion_calls, limitMultiFused = 5)

# filter fusion_calls to keep recurrent fusions from above sample.count and fusion_calls.summary
QCGeneFiltered_recFusion <- fusion_calls %>%
  dplyr::filter(FusionName %in% unique(c(sample.count$FusionName, fusion_calls.summary$FusionName)))

# filter QCGeneFiltered_recFusion to remove fusions found in more than 1 group and multifused gene per samples
QCGeneFiltered_recFusionUniq <- QCGeneFiltered_recFusion %>%
  dplyr::filter(!FusionName %in% group.count$FusionName) %>%
  dplyr::filter(!Gene1A %in% fusion_recurrent5_per_sample$GeneSymbol |
    !Gene2A %in% fusion_recurrent5_per_sample$GeneSymbol |
    !Gene1B %in% fusion_recurrent5_per_sample$GeneSymbol |
    !Gene2B %in% fusion_recurrent5_per_sample$GeneSymbol)
```


```{r}
putative_driver_fusions <- putative_driver_fusions %>%
  bind_rows(QCGeneFiltered_recFusionUniq[, colnames(putative_driver_fusions)]) %>%
  # filter fusions found in more than 4 histologies
  dplyr::filter(
    !FusionName %in% found_in_morethan_4$FusionName,
    # remove intergenic and intragenic fusion
    BreakpointLocation == "Genic",
    Fusion_Type == "in-frame"
  ) %>%
  as.data.frame()
```


```{r fig.width=6, fig.height=5}
# recurrently fused genes
plot_recurrent_genes(standardFusioncalls = putative_driver_fusions, groupby = "broad_histology", countID = "Kids_First_Participant_ID", plotn = 20)
```


```{r fig.width=6, fig.height=5}
# recurrent fusions
plot_recurrent_fusions(standardFusioncalls = putative_driver_fusions, groupby = "broad_histology", countID = "Kids_First_Participant_ID", plotn = 20)
```


```{r}
# PLot summary
p3 <- plot_summary(standardFusioncalls = putative_driver_fusions, groupby = "broad_histology", outputpdffile = "Summary.pdf")
```


