---
title: > 
  TCGA validation
author:
- name: Krutika Gaonkar
  affiliation: 
  - &id1 Center for Data-Driven Discovery in Biomedicine, Children’s Hospital of Philadelphia, USA
  email: gaonkark@chop.email.edu
- name: Federico Marini
  affiliation: 
  - &id2 Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz, Germany
  - Center for Thrombosis and Hemostasis (CTH), Mainz, Germany
  email: marinif@uni-mainz.de
date: "`r Sys.Date()`"
package: "`r paste('annoFuse', packageVersion('annoFuse'))"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{TCGA validation }
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{annoFuse}
  %\VignetteKeywords{Software, Transcription, Visualization, GeneExpression, Annotation,
    GeneFusionDetection, GUI, ReportWriting}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


```{r}
suppressPackageStartupMessages(library("readr"))
suppressPackageStartupMessages(library("ggplot"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("qdapRegex"))
```


# Introduction
To assess our filtering strategy, we analyzed a subset of samples from TCGA and compared fusions retained with annoFuse filtering and prioritization to those deemed the final call set in a previously published analysis by The Fusion Analysis Working Group.

A group of 160 samples were randomly selected BLCA(10), BRCA(11), CESC(5), COAD(11),  ESCA(5) , GBM(7), HNSC(10), KIRP(9), LGG(9), LIHC(9), LUAD(5), LUSC(11), OV(9), PAAD(8), PCPG(2), PRAD(14), SARC(6), SKCM(9), TGCT(6), THCA(4). 


```{r }
# manifest was obtained from CAVATICA Data browser TCGA files copied to working project
tcga_manifest <- read_tsv("../inst/extdata/sample_aliquot_tcga.tsv")

# Final fusion calls from https://www.sciencedirect.com/science/article/pii/S2211124718303954
final_fusion <- read_tsv("../inst/extdata/final_fusion_mmc2.txt") %>%
  # subset of final fusions in overlapping samples 
  dplyr::filter(
    Sample %in% tcga_manifest$aliquot_id ) %>% 
  as.data.frame() %>% 
  unique()

```

We ran [kf-rnaseq-workflow](https://github.com/kids-first/kf-rnaseq-workflow) to obtain fusion calls from STAR-Fusion and Arriba and rna quantification from RSEM . We then standardized STAR-Fusion and Arriba fusion calls using `fusion_standardization`. 


```{r }

# merged STAR-Fusion and arriba calls
tcga_arriba_df <- read_tsv("../inst/extdata/merged_arriba_tcga.tsv")
tcga_starfusion_df <-read_tsv("../inst/extdata/merged_starfusion_tcga.tsv")

# format
fomatted_starfusion<-annoFuse::fusion_standardization(fusion_calls = tcga_starfusion_df,caller = "STARFUSION",tumorID = tcga_starfusion_df$Sample)

# format
fomatted_arriba<-annoFuse::fusion_standardization(fusion_calls = tcga_arriba_df,caller = "ARRIBA",tumorID = tcga_arriba_df$Sample)


# Merge starfusion and arriba
formatted_calls<- rbind(fomatted_starfusion,fomatted_arriba)

```

Artifact and QC filtering was performed using `fusion_filtering_QC`. We filtered fusion using a wide range of spanningDelta : 10,20,30,40,50,100,150 and 200 

```{r}

merged_10<-annoFuse::fusion_filtering_QC(standardFusioncalls = formatted_calls,readthroughFilter = FALSE,readingFrameFilter = "in-frame|frameshift|other",artifactFilter = "GTEx_Recurrent|DGD_PARALOGS|Normal|BodyMap|ConjoinG",junctionReadCountFilter = 1, spanningFragCountFilter = 10)

merged_20<-annoFuse::fusion_filtering_QC(standardFusioncalls = formatted_calls,readthroughFilter = FALSE,readingFrameFilter = "in-frame|frameshift|other",artifactFilter = "GTEx_Recurrent|DGD_PARALOGS|Normal|BodyMap|ConjoinG",junctionReadCountFilter = 1, spanningFragCountFilter = 20)

merged_30<-annoFuse::fusion_filtering_QC(standardFusioncalls = formatted_calls,readthroughFilter = FALSE,readingFrameFilter = "in-frame|frameshift|other",artifactFilter = "GTEx_Recurrent|DGD_PARALOGS|Normal|BodyMap|ConjoinG",junctionReadCountFilter = 1, spanningFragCountFilter = 30)

merged_40<-annoFuse::fusion_filtering_QC(standardFusioncalls = formatted_calls,readthroughFilter = FALSE,readingFrameFilter = "in-frame|frameshift|other",artifactFilter = "GTEx_Recurrent|DGD_PARALOGS|Normal|BodyMap|ConjoinG",junctionReadCountFilter = 1, spanningFragCountFilter = 40)

merged_50<-annoFuse::fusion_filtering_QC(standardFusioncalls = formatted_calls,readthroughFilter = FALSE,readingFrameFilter = "in-frame|frameshift|other",artifactFilter = "GTEx_Recurrent|DGD_PARALOGS|Normal|BodyMap|ConjoinG",junctionReadCountFilter = 1, spanningFragCountFilter = 50)

merged_100<-annoFuse::fusion_filtering_QC(standardFusioncalls = formatted_calls,readthroughFilter = FALSE,readingFrameFilter = "in-frame|frameshift|other",artifactFilter = "GTEx_Recurrent|DGD_PARALOGS|Normal|BodyMap|ConjoinG",junctionReadCountFilter = 1, spanningFragCountFilter = 100)

merged_150<-annoFuse::fusion_filtering_QC(standardFusioncalls = formatted_calls,readthroughFilter = FALSE,readingFrameFilter = "in-frame|frameshift|other",artifactFilter = "GTEx_Recurrent|DGD_PARALOGS|Normal|BodyMap|ConjoinG",junctionReadCountFilter = 1, spanningFragCountFilter = 150)

merged_200<-annoFuse::fusion_filtering_QC(standardFusioncalls = formatted_calls,readthroughFilter = FALSE,readingFrameFilter = "in-frame|frameshift|other",artifactFilter = "GTEx_Recurrent|DGD_PARALOGS|Normal|BodyMap|ConjoinG",junctionReadCountFilter = 1, spanningFragCountFilter = 200)


```



```{r}
# get overlap with sample_id/aliquot_id with completed tasks
get_tcga_overlap<-function(x,tcga_manifest,final_fusion){
    
  samples_in_final_fusion <- x %>% 
    left_join(tcga_manifest,by=c("Sample"="sample_id")) %>%
    dplyr::filter(aliquot_id %in% final_fusion$Sample) %>%
    dplyr::select(aliquot_id,FusionName) %>% unique()
  
  final_fusion_subset <- final_fusion %>% dplyr::filter(Sample %in% samples_in_final_fusion$aliquot_id & Fusion %in% samples_in_final_fusion$FusionName) %>% nrow()
  return(final_fusion_subset)
  
}

```

 We then calculated overlaps of the subsetted published final fusion to our filtered fusion calls across all the cutoffs


```{r}
# get merged files for cutoff 10,20,30,40,50,100,150,200 with sensitivity 
length_final_fusion_subset <- final_fusion$Fusion %>% length() # 603
hist_df<-merged_10 %>%
  dplyr::mutate("cutoff"=10,"sensitivity"=get_tcga_overlap(merged_10,tcga_manifest ,final_fusion )/length_final_fusion_subset) %>%
  dplyr::bind_rows(merged_20 %>% 
                     dplyr::mutate("cutoff" = 20,"sensitivity"=get_tcga_overlap(merged_20,tcga_manifest ,final_fusion )/length_final_fusion_subset) ) %>%
  dplyr::bind_rows(merged_30 %>% 
                     dplyr::mutate("cutoff" = 30,"sensitivity"=get_tcga_overlap(merged_30,tcga_manifest ,final_fusion )/length_final_fusion_subset) ) %>%
  dplyr::bind_rows(merged_40 %>% 
                     dplyr::mutate("cutoff" = 40, "sensitivity"=get_tcga_overlap(merged_40,tcga_manifest ,final_fusion )/length_final_fusion_subset) ) %>%
  dplyr::bind_rows(merged_50 %>% 
                     dplyr::mutate("cutoff" = 50,"sensitivity"=get_tcga_overlap(merged_50,tcga_manifest ,final_fusion )/length_final_fusion_subset) ) %>%
  dplyr::bind_rows(merged_100 %>% 
                     dplyr::mutate("cutoff" = 100, "sensitivity"=get_tcga_overlap(merged_100,tcga_manifest ,final_fusion )/length_final_fusion_subset) ) %>%
  dplyr::bind_rows(merged_150 %>% 
                     dplyr::mutate("cutoff" = 150, "sensitivity"=get_tcga_overlap(merged_150,tcga_manifest ,final_fusion )/length_final_fusion_subset) ) %>%
  dplyr::bind_rows(merged_200 %>% 
                     dplyr::mutate("cutoff" = 200, "sensitivity"=get_tcga_overlap(merged_200,tcga_manifest ,final_fusion )/length_final_fusion_subset) ) %>%
  dplyr::mutate("change_spanning"=.$SpanningFragCount-.$JunctionReadCount)

hist_df <- as.data.frame(hist_df)
dat_text <- data.frame(
  label = c(paste("Mean",round(mean(hist_df[which(hist_df$cutoff==10),"change_spanning"]))), 
            paste("Mean",round(mean(hist_df[which(hist_df$cutoff==20),"change_spanning"]))),
            paste("Mean",round(mean(hist_df[which(hist_df$cutoff==30),"change_spanning"]))),
            paste("Mean",round(mean(hist_df[which(hist_df$cutoff==40),"change_spanning"]))), 
            paste("Mean",round(mean(hist_df[which(hist_df$cutoff==50),"change_spanning"]))), 
            paste("Mean",round(mean(hist_df[which(hist_df$cutoff==100),"change_spanning"]))), 
            paste("Mean",round(mean(hist_df[which(hist_df$cutoff==150),"change_spanning"]))),
            paste("Mean",round(mean(hist_df[which(hist_df$cutoff==200),"change_spanning"])))),
  cutoff   = c(10,20,30,40,50,100,150,200)
)


```

Here, we visualize the distribution of spanningDelta (spanningFragCount - JunctionReadCount) across fusions called from the TCG to assess a cutoff for spanningDelta

```{r fig.height= 4 }
ggplot(hist_df,aes(x=change_spanning))+geom_histogram(binwidth = 5) + 
  scale_x_continuous(limits = c(0,200)) + facet_wrap(.~cutoff) + 
  theme_publication(base_size = 20) +  
  geom_text(
  data    = dat_text,
  mapping = aes(x = 100, y = 1200, label = label),
  hjust   = -0.1,
  vjust   = -1
)+ theme(axis.title.x = element_text(size = 15),axis.title.y = element_text(size = 15))

```


We find that sensitivity reaches 96% at cutoff of 100 

```{r fig.height= 3, fig.width= 3}

extra_breaks <- c(0.7, 0.8,.85, 0.92, 0.96, 1.00)
breaks <- sort(c(extra_breaks, with(hist_df, pretty(range(sensitivity)))))

ggplot(hist_df,aes(x=cutoff,y=sensitivity))+
  geom_line() + ylab("sensitivity")+
  theme_publication(base_size = 20)+
  scale_y_continuous(
    breaks = breaks,
    limits = range(breaks)
  )+ theme(axis.title.x = element_text(size = 15),axis.title.y = element_text(size = 15))


```

